How economic, health and social factors affect life expectancy in developing countries?

### 2025-04-10 paper.Rmd
changed intro

Life expectancy is a fundamental indicator of a country's overall health and development. It reflects the effectiveness of healthcare systems, economic stability, and social progress. In developing countries healthcare resources are often constrained and economic disparities persist. Here, understanding the factors that influence life expectancy is critical for shaping effective policies. This study aims to answer the question: How economic, health and social factors influence life expectancy in developing countries, collectively and individually?

A multiple linear regression model is used, incorporating GDP per capita, alcohol consumption (categorical variable), schooling, infant mortality, health expenditure (as a percentage of a GDP per capita), and HIV deaths as explanatory variables. Existing literature provides strong evidence supporting the relationship between these factors and life expectancy. Miladinov (2020) examined the link between socioeconomic development and life expectancy in EU accession candidate countries, finding that higher GDP per capita and lower infant mortality rates were associated with increased life expectancy. Our study focuses on developing nations facing similar economic and healthcare challenges.

Adebayo et al. (2024) analyzed data in the United States, finding higher health expenditure and GDP per capita positively impact life expectancy while higher infant mortality reduces it. Our study extends this analysis to developing countries, allowing us to compare how these determinants function in different economic contexts. Our study provides a broader perspective on additional health risks that impact life expectancy in low-income nations.

A third study on OECD countries (Roffia et al., 2022) found that education and economic conditions played a critical role in increasing life expectancy. This supports our inclusion of schooling as a key predictor, as education directly influences health outcomes and economic mobility.

Life expectancy is influenced by a combination of factors, making a univariate approach insufficient. Multiple linear regression (MLR) is well-suited for this study as it quantifies the relationship between life expectancy (continuous response variable) and multiple predictors. MLR allows for simultaneous control of confounding variables, ensuring each predictor’s effect is estimated independently. The model provides interpretable coefficients allowing us to measure each socioeconomic and health factor’s impact on a developing country’s life expectancy. An interpretation of these results can help developing countries understand how to improve life expectancy.

>>>>

Life expectancy is a fundamental indicator of a country's overall health and development. It reflects the effectiveness of healthcare systems, economic stability, and social progress. In developing countries, healthcare resources are often constrained and economic disparities persist. Understanding the factors that influence life expectancy is critical for shaping effective policies. This study examines how economic, health, and social factors influence life expectancy in developing countries, collectively and individually.

A multiple linear regression model is used, incorporating alcohol consumption (categorical variable), schooling, infant mortality, health expenditure (as a percentage of a country’s GDP per capita), and HIV deaths as explanatory variables. We also included GDP per capita in the preliminary model but later removed it from the final set of predictors. Existing literature provides strong evidence supporting the relationship between these factors and life expectancy. Miladinov (2020) examined socioeconomic development and life expectancy in EU accession candidate countries, finding that higher GDP per capita and lower infant mortality were associated with increased life expectancy.

Our study focuses on developing nations facing similar economic and healthcare challenges. Adebayo et al. (2024) analyzed U.S. data, finding that higher health expenditure and GDP per capita positively impact life expectancy, while higher infant mortality reduces it. Our study extends this analysis to developing countries, comparing how these determinants function in different economic contexts. A third study on OECD countries (Roffia et al., 2022) found that education and economic conditions played a critical role in increasing life expectancy.

This supports schooling as a key predictor, as education directly influences health outcomes and economic mobility. To analyze these relationships, we employ multiple linear regression, allowing us to examine the relationship between life expectancy and multiple predictor variables simultaneously. Life expectancy is influenced by a combination of factors, making a univariate approach insufficient for capturing this complexity. This enables us to quantify the individual impact of each variable while controlling for others. This study seeks to generate actionable insights for policymakers to improve life expectancy in developing nations.
